% user_guide.tex — complete reference for every user-facing feature
\documentclass[11pt,a4paper]{article}
\input{Setup/Packages}
\input{Setup/Shortcuts}
\title{User Guide --- \cgspkg{} \cgsversion}
\author{Nicolas Flipo}
\date{\today}

\begin{document}
\maketitle

This guide is the authoritative reference for every user-facing command,
configuration option, document format, and error condition exposed by
\cgspkg{}.

% -----------------------------------------------------------------------
\section{CLI Overview}

The single entry-point is the \cgscli{} command.

\begin{lstlisting}[language=bash]
$ cgitsync <command> [options]
\end{lstlisting}

Running \cgscli{} without a command prints a help summary and exits with
code~0.

% -----------------------------------------------------------------------
\section{Commands}

\subsection{\texttt{validate}}

\begin{lstlisting}[language=bash]
$ cgitsync validate <spec.cgs>
\end{lstlisting}

Parses the \texttt{.cgs} file, builds the dependency-tree registry, and
prints a one-line summary for each repository node. Exits~0 on success,
non-zero on parse or validation error.

\textbf{Output fields:} \texttt{name}, \texttt{lifecycle\_state},
\texttt{complete}, \texttt{node\_count}.

\subsection{\texttt{tree}}

\begin{lstlisting}[language=bash]
$ cgitsync tree <spec.cgs>
\end{lstlisting}

Renders the dependency graph as a text tree. When a newer runtime snapshot is
registered for the given \texttt{.cgs} file, \cgscli{} loads that
\texttt{.gts} first so the tree reflects the current synchronised state.
Without a newer snapshot, it falls back to the declared \texttt{.cgs}
topology.

The command also accepts a direct \texttt{.gts} input:

\begin{lstlisting}[language=bash]
$ cgitsync tree <snapshot.gts>
\end{lstlisting}

\subsection{\texttt{describe}}

\begin{lstlisting}[language=bash]
$ cgitsync describe <snapshot.gts>
\end{lstlisting}

Reads a \texttt{.gts} snapshot and prints its contents as pretty-printed
JSON. Useful for inspecting saved tree state after a \texttt{clone} or
\texttt{freeze\_release} operation.

\subsection{\texttt{clone}}

\begin{lstlisting}[language=bash]
$ cgitsync clone <spec.cgs> [--target-dir DIR]
\end{lstlisting}

Clones the full repository tree declared by the \texttt{.cgs} file. The root
project directory defaults to \texttt{./<project-name>}. If that path already
exists and is not empty, \cgscli{} chooses the next available suffixed
directory such as \texttt{./ComplexGitSync-1}. When \texttt{--target-dir} is
provided, that path is used exactly.

For each repository, \cgscli{} tries the repo target branch first
(typically \texttt{default\_branch = "autoTest"}) and falls back to
\texttt{fallback\_branch} (typically \texttt{"main"}) when the preferred
branch does not exist on the remote. Nested \texttt{.cgs} files are discovered
after parent repositories are cloned, so explicit and automatic nested trees
are materialised during the same command.

On success, \texttt{clone} exits~0 and prints a READY summary including the
selected root path. It also writes a runtime snapshot to
\texttt{<project-root>/.cgitsync/state/<spec-name>.gts} and registers that
snapshot as the latest runtime state for the source \texttt{.cgs} file.

\subsection{\texttt{checkout} (Python API)}

Available via the Python API (CLI wiring planned):

\begin{lstlisting}[language=Python]
client.checkout("feature/my-branch")
# or with an explicit ref kind
from ComplexGitSync import RefKind
client.checkout("v1.2.0", ref_kind=RefKind.TAG)
\end{lstlisting}

Checks out a branch or tag across the whole tree, parent-first.
Requires a \texttt{READY} tree.  On success the tree remains \texttt{READY}
and a new \texttt{.gts} snapshot is written.

The operation follows three steps in order:

\begin{enumerate}
  \item \textbf{Propagate} — set the target ref on every registry entry
    (\texttt{propagate\_global\_branch}).
  \item \textbf{Create} — run \texttt{git branch} in every repo that does not
    yet have the branch locally (\texttt{create\_global\_branch}).
  \item \textbf{Checkout} — run \texttt{git checkout} in every repo,
    parent-first, and refresh each entry's ref and SHA.
\end{enumerate}

\subsection{\texttt{commit} (Python API)}

Available via the Python API (CLI wiring planned):

\begin{lstlisting}[language=Python]
# Stage all changes, then commit (default)
client.commit("feat: my change")

# Commit only already-staged changes
client.commit("feat: my change", stage_all=False)
\end{lstlisting}

Commits changes across the tree, leaf-first.  Repositories with no staged
changes after the optional \texttt{git add --all} step are silently skipped.
Requires a \texttt{READY} tree; the tree remains \texttt{READY} on success.

\subsection{\texttt{push} (Python API)}

Available via the Python API (CLI wiring planned):

\begin{lstlisting}[language=Python]
client.push()
\end{lstlisting}

Pushes all repositories to their configured remotes, leaf-first.  Each repo
is pushed to \texttt{entry.remote\_name} (defaulting to \texttt{"origin"}) on
its currently resolved branch.  Requires a \texttt{READY} tree; the tree
remains \texttt{READY} on success.

\subsection{\texttt{tag} (Python API)}

Available via the Python API (CLI wiring planned):

\begin{lstlisting}[language=Python]
client.tag("v1.2.3")
\end{lstlisting}

Creates and pushes the same tag across the full tree, leaf-first. Requires a
\texttt{READY} tree and keeps the tree \texttt{READY} on success.

\subsection{\texttt{freeze\_release} (Python API)}

Available via the Python API (CLI wiring planned):

\begin{lstlisting}[language=Python]
client.freeze_release(
    "release-2026.05",
    output_gts=".cgitsync/releases/release-2026.05.gts",
)
\end{lstlisting}

Performs a release freeze across all repos, leaf-first: stage/commit,
create+push the shared release tag, refresh SHA/state, and write a named
\texttt{.gts} snapshot. Requires \texttt{READY} and leaves the tree
\texttt{READY}.

\subsection{\texttt{launch\_release} (Python API)}

Available via the Python API (CLI wiring planned):

\begin{lstlisting}[language=Python]
client.launch_release(".cgitsync/releases/release-2026.05.gts")
\end{lstlisting}

Loads a release \texttt{.gts}, reconstructs the tree, performs required
clone/checkout actions, and must end in \texttt{READY} or fail explicitly.

\subsection{Planned Commands (not yet implemented)}

The following commands are defined in the public API contract but are not yet
implemented. Each returns exit code~2 with the message
\texttt{"not implemented yet"}.

\begin{itemize}
  \item \texttt{checkout} (CLI) — switch the whole tree to a named ref
    (Python API available; see above).
  \item \texttt{commit} (CLI) — stage and commit changes across the tree
    (Python API available; see above).
  \item \texttt{push} (CLI) — push all repos to their remotes
    (Python API available; see above).
  \item \texttt{restart} — re-synchronise a tree from an existing \texttt{.gts} snapshot.
  \item \texttt{tag} (CLI) — create a named tag across the tree
    (Python API available; see above).
  \item \texttt{freeze\_release} (CLI) — freeze a release and write a named
    \texttt{.gts} snapshot (Python API available; see above).
  \item \texttt{launch\_release} (CLI) — check out a previously frozen release
    (Python API available; see above).
\end{itemize}

% -----------------------------------------------------------------------
\section{Document Formats}

\subsection{Local Authoring Spec (\texttt{.cgs})}

A TOML file that describes the repository tree to manage. It is the only
hand-authored input; it is never modified at runtime.

\textbf{Top-level sections:}

\begin{itemize}
  \item \texttt{[document]} — mandatory; contains \texttt{format\_version}.
  \item \texttt{[project]} — mandatory; project-level metadata (see
        Table~\ref{tab:project-fields}).
  \item \texttt{[[repos]]} — one entry per managed repository
        (see Table~\ref{tab:repo-fields}).
\end{itemize}

\begin{table}[h]
\centering
\begin{tabular}{lll}
\hline
\textbf{Field} & \textbf{Type} & \textbf{Description} \\
\hline
\texttt{name}           & string  & Human-readable project label \\
\texttt{repo\_name}     & string  & Git repository slug \\
\texttt{default\_branch}& string  & Default branch (e.g.\ \texttt{main}) \\
\texttt{gitprovider}    & string  & Provider (\texttt{github}, \texttt{gitlab}, \texttt{custom}) \\
\texttt{gitprovider\_url}& string & Required for \texttt{custom} provider \\
\hline
\end{tabular}
\caption{Project-level fields in \texttt{.cgs}}
\label{tab:project-fields}
\end{table}

\begin{table}[h]
\centering
\begin{tabular}{lll}
\hline
\textbf{Field} & \textbf{Type} & \textbf{Description} \\
\hline
\texttt{gitprovider}         & string  & Provider override for this repo \\
\texttt{project\_owner\_name}& string  & GitHub owner / GitLab group \\
\texttt{project\_name}       & string  & Repository name on the provider \\
\texttt{relative\_path}      & string  & Path relative to project root \\
\texttt{gitprovider\_url}    & string  & Required for \texttt{custom} provider \\
\hline
\end{tabular}
\caption{Per-repository fields in \texttt{.cgs}}
\label{tab:repo-fields}
\end{table}

\subsection{Git Tree State Snapshot (\texttt{.gts})}

A TOML file generated automatically by bootstrapping and release operations.
It must never be hand-edited. It captures the exact commit SHAs and lifecycle
state of the tree at a point in time.

\textbf{Top-level sections:}
\begin{itemize}
  \item \texttt{[document]} — format version, \texttt{generated\_at},
        \texttt{command\_origin}.
  \item \texttt{[project]} — project name and \texttt{root\_absolute\_path}.
  \item \texttt{[tree\_state]} — \texttt{lifecycle\_state}, \texttt{is\_ready},
        \texttt{registry\_complete}.
  \item \texttt{[[repo\_state]]} — one entry per repo; includes
        \texttt{commit\_sha}, \texttt{sync\_state}, and ref information.
\end{itemize}

\subsection{Planning / GOC Documents (\texttt{.goc})}

Used for project identity and planning metadata. Supported in TOML, YAML, or
JSON. YAML support requires the optional \texttt{PyYAML} dependency.

% -----------------------------------------------------------------------
\section{Python API}

\cgspkg{} can be used as a library. The public API is exported from the
top-level package:

\begin{lstlisting}[language=Python]
from ComplexGitSync import (
    ComplexGitSyncClient,
    CgsDocument,
    GtsDocument,
    GocDocument,
    GitProvider,
    AccessProtocol,
    RepoAddress,
    GitRepo,
    GitTree,
    Orchestre,
    DependencyTreeRegistry,
)
\end{lstlisting}

\subsection{\texttt{ComplexGitSyncClient}}

High-level client that mirrors the CLI commands. Instantiate with a parsed
configuration path and call the equivalent method:

\begin{lstlisting}[language=Python]
from ComplexGitSync import ComplexGitSyncClient

client = ComplexGitSyncClient()
registry = client.load_cgs("project.cgs")
tree = client.format_project_tree()
clone_registry = client.clone_cgs("project.cgs", target_dir="./workspace/demo")
\end{lstlisting}

Once the tree is \texttt{READY}, the synchronisation methods are available:

\begin{lstlisting}[language=Python]
# Load an existing READY snapshot
client.load_gts(".cgitsync/state/project.gts")

# Check out a branch across all repos (parent-first)
client.checkout("feature/my-branch")
# or with a tag ref
from ComplexGitSync import RefKind
client.checkout("v1.0.0", ref_kind=RefKind.TAG)

# Commit across all repos (leaf-first), staging all changes by default
client.commit("feat: my change")
# or commit only already-staged changes
client.commit("feat: my change", stage_all=False)

# Push all repos to their remotes (leaf-first)
client.push()

# Create and push a shared tag (leaf-first)
client.tag("v1.2.3")

# Freeze release and emit a named snapshot
client.freeze_release("release-2026.05", output_gts=".cgitsync/releases/release-2026.05.gts")

# Launch from a frozen snapshot
client.launch_release(".cgitsync/releases/release-2026.05.gts")
\end{lstlisting}

\texttt{checkout}, \texttt{commit}, \texttt{push}, \texttt{tag}, and
\texttt{freeze\_release} require \texttt{READY} and leave the tree
\texttt{READY} on success. \texttt{launch\_release} loads a release
\texttt{.gts}, performs required clone/checkout actions, and must end in
\texttt{READY}. \texttt{checkout} and \texttt{freeze\_release} also write new
\texttt{.gts} snapshots.

\subsection{Document Classes}

Each document class exposes six format factories:

\begin{lstlisting}[language=Python]
doc = CgsDocument.from_toml(raw_bytes)
doc = CgsDocument.from_json(raw_str)
doc = CgsDocument.from_yaml(raw_str)   # requires PyYAML

serialised = doc.to_toml()
serialised = doc.to_json()
serialised = doc.to_yaml()             # requires PyYAML
\end{lstlisting}

\subsection{\texttt{RepoAddress}}

Composes and parses remote repository URLs:

\begin{lstlisting}[language=Python]
from ComplexGitSync import RepoAddress, GitProvider, AccessProtocol

addr = RepoAddress(
    gitprovider=GitProvider.GITHUB,
    owner_name="my-org",
    project_name="my-repo",
)
print(addr.to_ssh())    # git@github.com:my-org/my-repo.git
print(addr.to_https())  # https://github.com/my-org/my-repo.git
\end{lstlisting}

% -----------------------------------------------------------------------
\section{Configuration and Environment}

\cgspkg{} reads no environment variables and uses no global configuration
file for project topology. The logging and runtime-state subsystems use the
user state directory by default:

\begin{itemize}
  \item logs: \texttt{~/.local/state/ComplexGitSync/logs/}
  \item latest runtime-state mapping:
        \texttt{~/.local/state/ComplexGitSync/runtime/}
\end{itemize}

If \texttt{XDG\_STATE\_HOME} is set, these directories are rooted there
instead.

% -----------------------------------------------------------------------
\section{Error Reference}

\begin{description}
  \item[\texttt{ValueError}] Raised when a required field is missing or a
    field value is invalid (e.g.\ unknown \texttt{gitprovider},
    missing \texttt{gitprovider\_url} for a \texttt{custom} provider).
  \item[\texttt{GitSyncError}] Raised when clone-time Git operations fail,
    for example because the remote branch cannot be resolved or the target
    directory is already occupied.
  \item[Exit code 1] General CLI failure (parse error, validation error).
  \item[Exit code 2] Command exists but is not yet implemented.
\end{description}

% -----------------------------------------------------------------------
\section{Troubleshooting}

\begin{description}
  \item[``\texttt{gitprovider\_url} is required for custom provider'']
    Add a \texttt{gitprovider\_url} field to the offending
    \texttt{[[repos]]} entry in your \texttt{.cgs} file.
  \item[\texttt{validate} exits non-zero despite a correct-looking file]
    Check that \texttt{[document]} contains \texttt{format\_version = "1.0"}
    and that every \texttt{[[repos]]} block includes
    \texttt{project\_owner\_name} and \texttt{project\_name}.
  \item[\texttt{Clone destination already exists and is not empty}]
    Re-run \texttt{clone} with \texttt{--target-dir} to force a clean location,
    or let \cgscli{} choose a suffixed default directory.
  \item[\texttt{No cloneable branch found for <repo>}]
    Confirm that the remote exposes either the repo's
    \texttt{default\_branch} or its \texttt{fallback\_branch}.
  \item[PyYAML import error when using \texttt{.goc} files]
    Install the optional extra: \texttt{uv pip install "ComplexGitSync[yaml]"}.
\end{description}

\end{document}
