% getting_started.tex — step-by-step guide from installation to first use

\chapter{Getting Started}

This chapter guides a brand-new user from installation through the first
successful synchronisation of a nested Git repository tree.

The documented lifecycle is:

\begin{enumerate}
  \item \texttt{load(.cgs)} $\rightarrow$ \texttt{.gts LOADED}
  \item \texttt{expand(.cgs|.gts)} $\rightarrow$ \texttt{.gts PENDING}
  \item \texttt{validate(.cgs|.gts)} $\rightarrow$ \texttt{.gts READY}
  \item \texttt{clone(.cgs|.gts)}
  \item \texttt{checkout(.gts)}
  \item \texttt{add}
  \item \texttt{git(tree, "commit", msg)}
  \item \texttt{git(tree, "push")}
  \item \texttt{git(tree, "tag", name)}
  \item \texttt{freeze}
\end{enumerate}

% -----------------------------------------------------------------------
\section{Prerequisites}

\begin{itemize}
  \item Python~3.11 or later.
  \item Git~2.30 or later available on \texttt{PATH}.
  \item (Optional) PyYAML~6.x for \texttt{.yml} document support.
\end{itemize}

% -----------------------------------------------------------------------
\section{Installation}

Install \cgspkg{} from source with the Pixi-managed developer environment:

\begin{lstlisting}[language=bash]
$ git clone https://github.com/flipoyo/ComplexGitSync.git
$ cd ComplexGitSync
$ pixi install
\end{lstlisting}

Verify the installation:

\begin{lstlisting}[language=bash]
$ pixi run cgitsync --help
\end{lstlisting}

% -----------------------------------------------------------------------
\section{Creating a Project Spec (\texttt{.cgs})}

A \emph{local authoring spec} (\texttt{.cgs}) is a TOML file that describes
the repository tree you want to manage. For the self-standing
\texttt{ComplexGitSync} setup, create \texttt{complexgitsync.cgs}:

\begin{lstlisting}[language=toml]
[document]
format_version = "1.0"

[project]
name           = "ComplexGitSync"
default_branch = "autoTest"

[[repos]]
gitprovider        = "github"
project_owner_name = "flipoyo"
project_name       = "ComplexGitSync"
default_branch     = "autoTest"
fallback_branch    = "main"
relative_path      = "."

[[repos]]
gitprovider        = "github"
project_owner_name = "flipoyo"
project_name       = "DevSpec"
default_branch     = "autoTest"
fallback_branch    = "main"
relative_path      = "DevSpec"
nested_config      = "disabled"

[[repos]]
gitprovider        = "github"
project_owner_name = "flipoyo"
project_name       = "DocComplexGitSync"
default_branch     = "autoTest"
fallback_branch    = "main"
relative_path      = "docs"
nested_config      = "doccomplexgitsync.cgs"
\end{lstlisting}

In this topology, \texttt{docs/doccomplexgitsync.cgs} then declares
\texttt{DocSpec} (resolved as \texttt{docs/DocSpec}
from the repository root) as a nested documentation dependency.

This branch policy prefers the shared \texttt{autoTest} branch across the
whole tree. Each repo also declares \texttt{fallback\_branch = "main"}, so a
synchronisation run can stay on \texttt{autoTest} when that branch exists and
fall back to \texttt{main} when a repository has not created it yet.

% -----------------------------------------------------------------------
\section{Read / Expand / Verify}

Check that \cgscli{} can parse and resolve the spec without errors:

\begin{lstlisting}[language=bash]
$ cgitsync validate complexgitsync.cgs
\end{lstlisting}

A successful run exits with code~0 and prints the current tree summary, for
example \texttt{DECLARED ready=false complete=true}.

The canonical step-2 command is \texttt{expand}; \texttt{tree} is kept as a
backward-compatible alias.  The canonical step-3 command is \texttt{validate}.

% -----------------------------------------------------------------------
\section{Clone / Checkout}

Once the spec is valid, materialise the full project tree:

\begin{lstlisting}[language=bash]
$ cgitsync clone complexgitsync.cgs
\end{lstlisting}

By default, \cgscli{} clones the tree into \texttt{./ComplexGitSync}. If that
directory already exists and is not empty, the command chooses the next
available suffixed directory such as \texttt{./ComplexGitSync-1}. To force an
explicit destination, pass \texttt{--target-dir}.

On success, the command prints the explicit lifecycle workflow
(\texttt{load -> expand -> validate -> clone}), the command log file path
(\texttt{log\_file=...}), a minimal repo-only tree view, and a READY summary
including the chosen root path, for example
\texttt{READY ready=true complete=true root=/tmp/ComplexGitSync}.
The clone flow first tries each repo's \texttt{default\_branch} and falls back
to \texttt{fallback\_branch} when the preferred branch does not exist on the
remote. It also writes a runtime snapshot to
\texttt{<project-root>/.cgitsync/state/<spec-name>.gts} so later inspection
commands can reuse the synchronised state.

Once the tree is materialised, the lifecycle continues with checkout from the
recorded \texttt{.gts} state.

% -----------------------------------------------------------------------
\section{Inspecting and Printing Saved State}

Render the dependency graph as text:

\begin{lstlisting}[language=bash]
$ cgitsync tree complexgitsync.cgs
\end{lstlisting}

When a newer runtime snapshot exists for the same \texttt{.cgs} file,
\cgscli{} loads that \texttt{.gts} state automatically before rendering the
tree. This is why the same command reports a \texttt{READY} tree after a
successful \texttt{clone} instead of returning to the purely declared
\texttt{DECLARED}/\texttt{PENDING} view.

The remaining planned user-facing work is to expose a dedicated
\texttt{print(.gts)} command for saved-state inspection.

% -----------------------------------------------------------------------
\section{Preparing an Orchestration Plan (\texttt{.goc})}

A \texttt{.goc} file records a repeatable command sequence for a project.
For the self-standing setup, \texttt{deploy.goc} keeps the tree on
\texttt{autoTest} when possible:

\begin{lstlisting}[language=toml]
[[actions]]
command = "checkout"
[actions.args]
ref      = "autoTest"
ref_type = "branch"
\end{lstlisting}

Combined with the per-repo \texttt{fallback\_branch = "main"} settings from
the \texttt{.cgs} file, the same orchestration plan can be used on trees where
some repositories already expose \texttt{autoTest} and others still only offer
\texttt{main}. In the current bootstrap release, the \texttt{.goc} file is the
machine-readable reference for that flow; direct CLI execution of orchestration
plans is introduced later as \texttt{orchestrate(.goc)}.

% -----------------------------------------------------------------------
\section{Log Files}

Every CLI command writes a timestamped log file. On Linux the default location
is \texttt{~/.local/state/ComplexGitSync/logs/}; if
\texttt{XDG\_STATE\_HOME} is set, the logs are written under that state
directory instead. The logs record command start and end, tree and repo state
transitions, fallback decisions, nested \texttt{.cgs} discovery, and
\texttt{.gts} loads and writes. The CLI prints the resolved file path in
command output as \texttt{log\_file=<path>}.

% -----------------------------------------------------------------------
\section{Next Steps}

Once the tree is cloned you can proceed to:
\begin{itemize}
  \item \textbf{checkout} (Python API) — switch the whole tree to a branch or tag,
    parent-first.
  \item \textbf{add} / \textbf{commit} (Python API) — stage and commit changes
    across the tree, leaf-first.
  \item \textbf{push} (Python API) — push every repository to its configured
    remote, leaf-first, and refresh stored hashes.
  \item \textbf{tag} / \textbf{freeze} (Python API) — update recorded tag state
    and emit the next saved \texttt{.gts}.
\end{itemize}

All of these synchronisation operations require a \texttt{READY} tree and leave it
\texttt{READY} after a successful run.  See Chapter~\ref{chap:user-guide} for
complete command and API documentation.
