% user_guide.tex — complete reference for every user-facing feature
\chapter{User Guide}
\label{chap:user-guide}

This chapter is the authoritative reference for every user-facing command,
configuration option, document format, and error condition exposed by
\cgspkg{}.

% -----------------------------------------------------------------------
\section{CLI Overview}

The single entry-point is the \cgscli{} command.

\begin{lstlisting}[language=bash]
$ cgitsync <command> [options]
\end{lstlisting}

Running \cgscli{} without a command prints a help summary and exits with
code~0.

% -----------------------------------------------------------------------
\section{Lifecycle Contract}

This guide follows the simplified canonical lifecycle:

\begin{enumerate}
  \item \texttt{initialise(.cgs)} $\rightarrow$ clone all repos $\rightarrow$
        \texttt{READY}  \textit{(new project)}\\
        \texttt{initialise(.gts)} $\rightarrow$ restore from snapshot $\rightarrow$
        \texttt{READY}  \textit{(existing project)}
  \item \texttt{pull(.cgs|.gts)} $\rightarrow$ resync an existing tree
  \item \texttt{checkout(.gts)} / \texttt{add} / \texttt{commit} / \texttt{push} / \texttt{tag}
        $\rightarrow$ git operations on the tree
  \item \texttt{freeze} $\rightarrow$ emit the next \texttt{.gts} id
\end{enumerate}

Python API power users may also call \texttt{load(.cgs|.gts)} directly:
\texttt{.gts} sources are loaded directly; \texttt{.cgs} sources run the
internal expand/validate pipeline.

Compatibility aliases: \texttt{read} $\rightarrow$ \texttt{load},
\texttt{describe} $\rightarrow$ \texttt{print}, \texttt{restart} $\rightarrow$
\texttt{pull}, \texttt{freeze-release} $\rightarrow$ \texttt{freeze}.

The remaining planned work is \texttt{orchestrate(.goc)} and project-local
\texttt{.lgr} register management.

% -----------------------------------------------------------------------
\section{Commands}

\subsection{\texttt{initialise}}

\begin{lstlisting}[language=bash]
$ cgitsync initialise <spec.cgs> [--target-dir DIR]
$ cgitsync initialise <snapshot.gts>
\end{lstlisting}

Unified initialisation entry point.  Dispatches on file extension:

\begin{itemize}
  \item \texttt{.cgs} source (clone mode): clones the full repository tree
    (see \texttt{clone} below). The optional \texttt{--target-dir} controls
    where the project root is created.  On success, ends in \texttt{READY}.
  \item \texttt{.gts} source (restore mode): loads a saved snapshot directly.
    On success, ends in the lifecycle state recorded in the snapshot.
\end{itemize}

Both paths end in a \texttt{READY} tree or fail explicitly.

\subsection{\texttt{pull}}

\begin{lstlisting}[language=bash]
$ cgitsync pull <spec.cgs|snapshot.gts>
\end{lstlisting}

Resynchronises the project tree.  \texttt{.cgs} sources follow the
\texttt{restart} flow (load, discover, checkout current branch, READY), while
\texttt{.gts} sources follow saved-state restore semantics equivalent to
\texttt{launch-release}.  On success, the tree ends in \texttt{READY}.

\subsection{\texttt{clone}}

\begin{lstlisting}[language=bash]
$ cgitsync clone <spec.cgs> [--target-dir DIR]
\end{lstlisting}

Clones the full repository tree declared by the \texttt{.cgs} file. The root
project directory defaults to \texttt{./<project-name>}. If that path already
exists and is not empty, \cgscli{} chooses the next available suffixed
directory such as \texttt{./ComplexGitSync-1}. When \texttt{--target-dir} is
provided, that path is used exactly.

For each repository, \cgscli{} tries the repo target branch first
(typically \texttt{default\_branch = "autoTest"}) and falls back to
\texttt{fallback\_branch} (typically \texttt{"main"}) when the preferred
branch does not exist on the remote. Nested \texttt{.cgs} files are discovered
after parent repositories are cloned, so explicit and automatic nested trees
are materialised during the same command.

On success, \texttt{clone} exits~0 and prints a READY summary including the
selected root path. It also writes a runtime snapshot to
\texttt{<project-root>/.cgitsync/state/<spec-name>.gts} and registers that
snapshot as the latest runtime state for the source \texttt{.cgs} file.

\subsection{\texttt{checkout}}

\begin{lstlisting}[language=bash]
$ cgitsync checkout <branch> --gts <snapshot.gts> [--ref-kind branch|tag]
\end{lstlisting}

Checks out a branch or tag across the whole tree, parent-first.
Loads the READY registry from \texttt{--gts}, then propagates the ref,
creates missing local branches, and runs \texttt{git checkout} in every repo.
On success the tree remains \texttt{READY} and a new \texttt{.gts} snapshot is written.

Also available via the Python API directly:

\begin{lstlisting}[language=Python]
client.checkout("feature/my-branch")
# or with an explicit ref kind
from ComplexGitSync import RefKind
client.checkout("v1.2.0", ref_kind=RefKind.TAG)
\end{lstlisting}

\subsection{\texttt{add}}

\begin{lstlisting}[language=bash]
$ cgitsync add --gts <snapshot.gts>
\end{lstlisting}

Stages all changes across the tree (\texttt{git add --all}), leaf-first.
Requires \texttt{READY}; tree stays \texttt{READY}.

\subsection{\texttt{commit}}

\begin{lstlisting}[language=bash]
$ cgitsync commit <message> --gts <snapshot.gts> [--no-stage]
\end{lstlisting}

Commits changes across the tree, leaf-first.  Loads the READY registry from
\texttt{--gts}.  Repositories with no staged changes after the optional
\texttt{git add --all} step (suppressed by \texttt{--no-stage}) are silently
skipped. Requires a \texttt{READY} tree; the tree remains \texttt{READY} on success.

The commit message conventionally ends with \texttt{CGS\#VERSION}.

Also available via the unified \texttt{git()} Python API:

\begin{lstlisting}[language=Python]
registry = client.get_dependency_registry()
# Stage all changes, then commit (default)
client.git(registry, "commit", "feat: my change CGS#1")

# Direct method: commit only already-staged changes
client.commit("feat: my change", stage_all=False)
\end{lstlisting}

\subsection{\texttt{push}}

\begin{lstlisting}[language=bash]
$ cgitsync push --gts <snapshot.gts>
\end{lstlisting}

Pushes all repositories to their configured remotes, leaf-first.  Loads the
READY registry from \texttt{--gts}.  Each repo is pushed to
\texttt{entry.remote\_name} (defaulting to \texttt{"origin"}) on its currently
resolved branch.  Refreshes the stored hash in \texttt{GitTree}.  Requires a
\texttt{READY} tree; the tree remains \texttt{READY} on success.

Also available via the unified \texttt{git()} Python API:

\begin{lstlisting}[language=Python]
registry = client.get_dependency_registry()
client.git(registry, "push")
\end{lstlisting}

\subsection{\texttt{tag}}

\begin{lstlisting}[language=bash]
$ cgitsync tag <name> --gts <snapshot.gts>
\end{lstlisting}

Creates and pushes the same tag across the full tree, leaf-first.  Loads the
READY registry from \texttt{--gts}.  Refreshes the stored tag in
\texttt{GitTree}.  Requires a \texttt{READY} tree and keeps the tree
\texttt{READY} on success.

Also available via the unified \texttt{git()} Python API:

\begin{lstlisting}[language=Python]
registry = client.get_dependency_registry()
client.git(registry, "tag", "v1.2.3")
\end{lstlisting}

\subsection{\texttt{freeze}}

\begin{lstlisting}[language=bash]
$ cgitsync freeze <name> --gts <snapshot.gts>
\end{lstlisting}

Performs a release freeze across all repos, leaf-first: stage/commit,
create+push the shared release tag, refresh SHA/state, and write a
\texttt{.gts} snapshot. Loads the READY registry from \texttt{--gts}.
Requires \texttt{READY} and leaves the tree \texttt{READY}.

Also available via the Python API:

\begin{lstlisting}[language=Python]
client.freeze(
    "release-2026.05",
    output_gts=".cgitsync/releases/release-2026.05.gts",
)
\end{lstlisting}

\texttt{freeze-release} and \texttt{freeze-state} remain available as
compatibility aliases for release and internal-state freeze operations
respectively.

\subsection{\texttt{print}}

\begin{lstlisting}[language=bash]
$ cgitsync print <spec.cgs|snapshot.gts>
\end{lstlisting}

Prints a lifecycle summary as pretty JSON. For \texttt{.gts} input, it reports
the recorded snapshot state. For \texttt{.cgs} input, it reports the current
declared/runtime summary.

\subsection{\texttt{describe}}

\begin{lstlisting}[language=bash]
$ cgitsync describe <snapshot.gts>
\end{lstlisting}

Reads a \texttt{.gts} snapshot and prints its contents as pretty-printed
JSON. This is a compatibility alias of \texttt{print}; both commands share the
same implementation.

\subsection{\texttt{restart}}

\begin{lstlisting}[language=bash]
$ cgitsync restart <spec.cgs>
\end{lstlisting}

Resynchronizes an already-cloned project tree. Compatibility alias for
\texttt{pull} with a \texttt{.cgs} source.

\subsection{\texttt{launch-release}}

\begin{lstlisting}[language=bash]
$ cgitsync launch-release <snapshot.gts>
\end{lstlisting}

Loads a release \texttt{.gts}, reconstructs the tree, performs required
clone/checkout actions, and must end in \texttt{READY} or fail explicitly.

Also available via the Python API:

\begin{lstlisting}[language=Python]
client.launch_release(".cgitsync/releases/release-2026.05.gts")
\end{lstlisting}

\subsection{Planned Commands (not yet implemented)}

The following commands are defined in the public API contract but are not yet
implemented. Each returns exit code~2 with the message
\texttt{"not implemented yet"}.

\begin{itemize}
  \item \texttt{write-gts} — write a \texttt{.gts} snapshot explicitly
    from a loaded \texttt{.cgs} or runtime state.
  \item \texttt{status} — summarise tree readiness and per-repo sync state.
  \item \texttt{orchestrate(.goc)} — planned execution surface for
    orchestration plans.
\end{itemize}

% -----------------------------------------------------------------------
\section{Document Formats}

\texttt{.cgs} means \textbf{ComplexGitSync} specification, \texttt{.gts} means
\textbf{GitTreeState} snapshot, \texttt{.goc} means
\textbf{GitOrchestrationCommand} plan, and \texttt{.lgr} means
\textbf{Local Git Register}.

\subsection{Local Authoring Spec (\texttt{.cgs})}

A TOML file that describes the repository tree to manage. It is the only
hand-authored input; it is never modified at runtime.

\textbf{Top-level sections:}

\begin{itemize}
  \item \texttt{[document]} — mandatory; contains \texttt{format\_version}.
  \item \texttt{[project]} — mandatory; project-level metadata (see
        Table~\ref{tab:project-fields}).
  \item \texttt{[[repos]]} — one entry per managed repository
        (see Table~\ref{tab:repo-fields}).
\end{itemize}

\begin{table}[h]
\centering
\begin{tabular}{lll}
\hline
\textbf{Field} & \textbf{Type} & \textbf{Description} \\
\hline
\texttt{name}           & string  & Human-readable project label \\
\texttt{repo\_name}     & string  & Git repository slug \\
\texttt{default\_branch}& string  & Default branch (e.g.\ \texttt{main}) \\
\texttt{gitprovider}    & string  & Provider (\texttt{github}, \texttt{gitlab}, \texttt{custom}) \\
\texttt{gitprovider\_url}& string & Required for \texttt{custom} provider \\
\hline
\end{tabular}
\caption{Project-level fields in \texttt{.cgs}}
\label{tab:project-fields}
\end{table}

\begin{table}[h]
\centering
\begin{tabular}{lll}
\hline
\textbf{Field} & \textbf{Type} & \textbf{Description} \\
\hline
\texttt{gitprovider}         & string  & Provider override for this repo \\
\texttt{project\_owner\_name}& string  & GitHub owner / GitLab group \\
\texttt{project\_name}       & string  & Repository name on the provider \\
\texttt{relative\_path}      & string  & Path relative to project root \\
\texttt{gitprovider\_url}    & string  & Required for \texttt{custom} provider \\
\hline
\end{tabular}
\caption{Per-repository fields in \texttt{.cgs}}
\label{tab:repo-fields}
\end{table}

\subsection{Git Tree State Snapshot (\texttt{.gts})}

A TOML file generated automatically by bootstrapping and release operations.
It must never be hand-edited. It captures the exact commit SHAs and lifecycle
state of the tree at a point in time.

\textbf{Top-level sections:}
\begin{itemize}
  \item \texttt{[document]} — format version, \texttt{generated\_at},
        \texttt{command\_origin}.
  \item \texttt{[project]} — project name and \texttt{root\_absolute\_path}.
  \item \texttt{[tree\_state]} — \texttt{lifecycle\_state}, \texttt{is\_ready},
        \texttt{registry\_complete}.
  \item \texttt{[[repo\_state]]} — one entry per repo; includes
        \texttt{commit\_sha}, \texttt{sync\_state}, and ref information.
\end{itemize}

\subsection{Planning / GOC Documents (\texttt{.goc})}

Used for project identity and planning metadata. Supported in TOML, YAML, or
JSON. YAML support requires the optional \texttt{PyYAML} dependency.

% -----------------------------------------------------------------------
\section{Python API}

\cgspkg{} can be used as a library. The public API is exported from the
top-level package:

\begin{lstlisting}[language=Python]
from ComplexGitSync import (
    ComplexGitSyncClient,
    CgsDocument,
    GtsDocument,
    GocDocument,
    GitProvider,
    AccessProtocol,
    RepoAddress,
    GitRepo,
    GitTree,
    Orchestre,
    DependencyTreeRegistry,
)
\end{lstlisting}

\subsection{\texttt{ComplexGitSyncClient}}

High-level client that mirrors the CLI commands. Instantiate with a parsed
configuration path and call the equivalent method:

\begin{lstlisting}[language=Python]
from ComplexGitSync import ComplexGitSyncClient

client = ComplexGitSyncClient()
registry = client.read("project.cgs")
tree_state = client.validate("project.cgs")
tree = client.format_project_tree()
clone_registry = client.clone("project.cgs", target_dir="./workspace/demo")
\end{lstlisting}

Once the tree is \texttt{READY}, the synchronisation methods are available:

\begin{lstlisting}[language=Python]
# Load an existing READY snapshot
client.load_gts(".cgitsync/state/project.gts")

# Check out a branch across all repos (parent-first)
client.checkout("feature/my-branch")
# or with a tag ref
from ComplexGitSync import RefKind
client.checkout("v1.0.0", ref_kind=RefKind.TAG)

# Commit across all repos (leaf-first), staging all changes by default
client.commit("feat: my change")
# or commit only already-staged changes
client.commit("feat: my change", stage_all=False)

# Push all repos to their remotes (leaf-first)
client.push()

# Create and push a shared tag (leaf-first)
client.tag("v1.2.3")

# Freeze and emit a named snapshot
client.freeze("release-2026.05", output_gts=".cgitsync/releases/release-2026.05.gts")

\end{lstlisting}

\texttt{checkout}, \texttt{commit}, \texttt{push}, \texttt{tag}, and
\texttt{freeze} require \texttt{READY} and leave the tree
\texttt{READY} on success. \texttt{checkout} and \texttt{freeze} also write new
\texttt{.gts} snapshots. If you need to restore a saved \texttt{.gts} directly,
see the compatibility helper documented in the \texttt{launch-release}
subsection above.

\subsection{Document Classes}

Each document class exposes six format factories:

\begin{lstlisting}[language=Python]
doc = CgsDocument.from_toml(raw_bytes)
doc = CgsDocument.from_json(raw_str)
doc = CgsDocument.from_yaml(raw_str)   # requires PyYAML

serialised = doc.to_toml()
serialised = doc.to_json()
serialised = doc.to_yaml()             # requires PyYAML
\end{lstlisting}

\subsection{\texttt{RepoAddress}}

Composes and parses remote repository URLs:

\begin{lstlisting}[language=Python]
from ComplexGitSync import RepoAddress, GitProvider, AccessProtocol

addr = RepoAddress(
    gitprovider=GitProvider.GITHUB,
    owner_name="my-org",
    project_name="my-repo",
)
print(addr.to_ssh())    # git@github.com:my-org/my-repo.git
print(addr.to_https())  # https://github.com/my-org/my-repo.git
\end{lstlisting}

% -----------------------------------------------------------------------
\section{Configuration and Environment}

\cgspkg{} reads no environment variables and uses no global configuration
file for project topology. The logging and runtime-state subsystems use the
user state directory by default:

\begin{itemize}
  \item logs: \texttt{~/.local/state/ComplexGitSync/logs/}
  \item latest runtime-state mapping:
        \texttt{~/.local/state/ComplexGitSync/runtime/}
\end{itemize}

If \texttt{XDG\_STATE\_HOME} is set, these directories are rooted there
instead.

% -----------------------------------------------------------------------
\section{Error Reference}

\begin{description}
  \item[\texttt{ValueError}] Raised when a required field is missing or a
    field value is invalid (e.g.\ unknown \texttt{gitprovider},
    missing \texttt{gitprovider\_url} for a \texttt{custom} provider).
  \item[\texttt{GitSyncError}] Raised when clone-time Git operations fail,
    for example because the remote branch cannot be resolved or the target
    directory is already occupied.
  \item[Exit code 1] General CLI failure (parse error, validation error).
  \item[Exit code 2] Command exists but is not yet implemented.
\end{description}

% -----------------------------------------------------------------------
\section{Troubleshooting}

\begin{description}
  \item[``\texttt{gitprovider\_url} is required for custom provider'']
    Add a \texttt{gitprovider\_url} field to the offending
    \texttt{[[repos]]} entry in your \texttt{.cgs} file.
  \item[\texttt{validate} exits non-zero despite a correct-looking file]
    Check that \texttt{[document]} contains \texttt{format\_version = "1.0"}
    and that every \texttt{[[repos]]} block includes
    \texttt{project\_owner\_name} and \texttt{project\_name}.
  \item[\texttt{Clone destination already exists and is not empty}]
    Re-run \texttt{clone} with \texttt{--target-dir} to force a clean location,
    or let \cgscli{} choose a suffixed default directory.
  \item[\texttt{No cloneable branch found for <repo>}]
    Confirm that the remote exposes either the repo's
    \texttt{default\_branch} or its \texttt{fallback\_branch}.
  \item[PyYAML import error when using \texttt{.goc} files]
    Add PyYAML in the Pixi environment: \texttt{pixi add pyyaml}.
\end{description}
